\include{zLaTex-Preamble}
\usepackage{fixltx2e}

\newcommand{\bib}{

\begin{thebibliography}{9}

\bibitem{beal}
  Beal, Richard H. 
  ``Gleanings from Hittite Oracle Questions on Religion, Society, Psychology and Decision Making.'' 
  Silvia Anatolica:Anatolian Studies Presented to Maciej Popko (2002): 11-37.

\bibitem{george}
  George, Andrew. 
  \emph{The Epic of Gilgamesh: The Babylonian Epic Poem and Other Texts in Arcadian and Sumerian}. 
  Penguin, 2002.
  Illustrated, reprint, reissue. 
  228 pages.

\bibitem{hoffner}
  Hoffner, Harry A., Jr.
  \emph{Hittite Myths}
  Writings from the Ancient World 2. 
  Atlanta: Scholars Press. 1990.

\bibitem{johnston}
  Johnston, S. I. 
  \emph{Religions of the Ancient World. A Guide.}
  Cambridge. Ma. 
  Harvard Univ. Press, 2004.

\bibitem{sasson}
  Sasson, Jack M. 
  \emph{Civilizations of the ancient Near East}. 
  Vol. 3. 
  New York: Scribner, 1995.

\bibitem{singer}
  Singer, Itamar.  
  \emph{Hittite prayers}.
  edited by Harry A. Hoffner, Jr.  
  ISBN 1-58983-032-6.

\bibitem{kramer}
  Kramer, Samuel Noah.
  \emph{From the Poetry of Sumer: Creation, Glorification, Adoration}
  University of California Press. 
  Berkely and Los Angeles, California.
  1979.
  ISBN 0-250-03703-0.

\bibitem{van_death}
 Bremer, J.M., van den Hout, Th.P.J., Peters, R.
 \emph{Death as a Privilege. The Hittite Royal Funerary Ritual} 
 Hidden Futures. Death and Immortality in Ancient Egypt, Anatolia, the Classical, Biblical and Arabic-Islamic World 
 Amsterdam. 37-75.
 1994.

\bibitem{van}
  van den Hout, Th PJ. 
  ``3. Biography and Autobiography.'' 
  \emph{The Context of Scripture: Canonical compositions from the biblical world 1}
  1997.

\end{thebibliography}
}


\title{\papertitle}
\author{Joshua Miller}
\date{March 14, 2014}

\linespread{2}

\begin{document}

\maketitle
\newpage

As the first literate inhabitants of
Mesopotamia \textsuperscript{\cite[p.20]{george}}, the Sumerians
provide an interesting point of comparison for successive
civilizations and the literary representations of their deities.  By
comparing the written mythology, poems, instruction texts and rituals
with those of the Hittites, we can gain insight to the relative views
of divinity between cultures.  Although the literature of the Hittites
represents their gods in a similar manner to those of the Sumerians,
the Hittite gods are viewed with more pragmatism than their Sumerian
counterparts.

Both Hittite and Sumerian mythologies present their gods as
anthropomorphic beings, with very human desires and emotions.
%% ~~~
A Sumerian example of a god exhibiting such human behaviors can be
found in the myth of Ishtar and the Bull of Heaven. The story begins
as the goddess Ishtar proposes marriage to Gilgamesh: ``let me be your
bride and you shall be my
husband'' \textsuperscript{\cite[p.102]{george}}. To this, Gilgamesh
responds that he ``will not. How would it go with me?  Your lovers
have found you like a brazier which smoulders in the cold, a backdoor
which keeps out neither squall of wind nor storm ... Which of your
lovers did you love for
ever?'' \textsuperscript{\cite[p.103]{george}}.
%% ~~~
This direct insult incites Ishtar to seek a revenge so saturated with
emotion that it is impossible to deny the human nature of the goddess.
%% ~~~
She pleas for permission to release the Bull of Heaven unto Uruk, and
in turn kills hundreds of innocent
people \textsuperscript{\cite[p.103]{george}}. Ishtar's actions can be
characterized as human and reactionary and not part of a divinely just
moral scheme, demonstrating the extent to which the Sumerian gods are
depicted almost as supernatural humans.
%% ~~~
In the Hittite disappearing deity myth, we see a similar human nature
in the gods.  Although many versions of this story forgo mention of
the cause of anger, each one involves an enraged god's disappearance
and the parallel failure of agriculture, fertility, and water sources.
%% ~~~
In the Telipinu version, for example, ``Telipinu too went away and
removed grain, animal fecundity, luxuriance, growth, and
abundance'' \textsuperscript{\cite[p.14]{hoffner}} and the other
Hittite gods became worried. The world remained dessicate of resources
until wax was applied to a bee sting, soothing the delinquent god's
anger.
%% ~~~
In both of the myths presented above, we see a shared anthropotheistic
representation of the deities between cultures. Moreover, these
stories serve to illustrate that both the Sumerian and Hittite deities
were representable in a physical form (the ability of the bee to sting
Telipinu, and the carnal request for a marriage to Isthar) and that
the pantheon was physitheistic. 
%% ~~~
Further evidence of this can be found in the mythos of the Sun God,
the Cow, and the Fisherman, as well as the tale of Appu and his
sons. The sun god not only attains a physical form when appearing to
Appu, ``as he changed himself into a young man, came to him, and
questioned him''\textsuperscript{\cite[p.83]{hoffner}}, but again
``[became] a young man, came down from the sky'' for a rendezvous with
a bovine lover\textsuperscript{\cite[p.85]{hoffner}}.
%% ~~~
As in the Sumerian myth involving Ishtar, we see that the Hittite gods
even capable of entering into corporeal engagements.  It should also
be noted that the representations of the gods were not solely
anthropomorphic, but allowed for ``zoomorphic [representations] such
as Alaca H\"uy\"uk .. where the storm-god is represented as a
bull'' \textsuperscript{\cite[p.1990]{sasson}}. This shared
anthropopathy is hardly surprising as it was normative of religions
during the time period.

In both Hittite and Sumerian literature, the king is portrayed as an
intermediary between the realm of the gods and of the mortals.  A
Sumerian king was granted the ability to ``journey all alone to
receive blessings from various gods. Shulgi, for example ... presents
offerings to Nanna and is further extolled and
blessed'' \textsuperscript{\cite[p.63]{Kramer}}.  In a similar manner,
the Hittite kings were ``an intermediary between god and man,
representative of the Sungod on earth. Although close to the gods, he
was not considered divine himself, but turned into a god at
death'' \textsuperscript{\cite[p.38]{van_death}}. We see that both
cultunres present the divine as approachable beings.  It is
interesting to note that Hittite literature commonly mentions the
transformation of human king to a god, whereas Sumerian literature is
absent of such rebirths. In Hittite texts, i.e. prayers, edicts, etc.,
it was common to say that a king ``became
god'' \textsuperscript{\cite[p.75,95,98]
{singer},\cite[p.196]{van}[etc.]} upon his death.
Furthermore, this acquired divinity may have been extended to other
members of the royal family.  Any questionability of actual divine
state of deceased kings, as opposed to being a simple formulaic
epithet devoid of substantial belief that kings became gods, is
quelled by other passages in the Hittite corpus. The ``intricate royal
funerary ritual lasting for two weeks and heavily laden with
symbolism'' \textsuperscript{\cite[p.38]{van_death}} as well as the
cult worship to ``statues of deceased kings [that] had been erected in
temples and received
offerings'' \textsuperscript{\cite[p.45]{van_death}} both serve this
purpose.
%% ~~~
In stark contrast however, every Sumerian mortal was consigned to an
existence in the morbid Netherworld after death, regardless of rank or
regality.
%% ~~~
Even though the legend of Gilgamesh curiously explains that ``Two
thirds they [the gods] made him [Gilgamesh], god and one third
man,''  \textsuperscript{\cite[p77]{george}} his ascension to full divinity is
nevertheless an impossibility.
%% ~~~
After the death of his companion, Enkidu, Gilgamesh begins to fear
death and takes on a quest for immortality because even ``the great king is
after all an ordinary mortal.'' \textsuperscript{\cite[p.47]{george}}. 
%% ~~~
While the Sumerian mythology presents gods in a separate sphere of
existence from humans, the Hittite literature blurs the distinction. 
%% ~~~
It would seem that the Hittites allowed for the ``Gods ascended from
kingship model'' while the Sumerian's theologians have ``summed it all
up with the statement 'Kingship descended from
heaven''' \textsuperscript{\cite[p.50]{kramer}}.
%% ~~
This is not to say that the Hittite gods were any less powerful.  The
ample prayers, festivals, and instruction texts clearly demonstrate
their vigor.  It simply means that the literature suggests a
connotation around Hittite divinity that differs from Sumerian texts;
the Hittite gods were anthropomorphized to a larger extent than the
Sumerian gods.

This difference can seen within a spectrum of pragmatism: adoration
versus the function of the gods as a utility.  To a large extent,
every ancient Near Eastern civilization found a great purpose in
religion as an explanation and remedy for worldly problems. Sumerian
literature presents divinity in terms of utility accompanied with
adoration.
%% ~~~ 
The motif of Sumerian adoration can be found in poems such as the
following glorification from Enki and the World Order:
\begin{quote}
You, lord Enlil, who are lord, god, and king, //
Who are the judge and the decision-maker of the universe, //
Your noble word is as weighty as heaven, you know no opposition //
... //
'O father Enki, praise! \textsuperscript{\cite[p.44]{kramer}}
\end{quote}
This poem serves the informative purpose of describing Enki's creation
of the world by means of praise and adoration.  Even Sumer's ``female
deities were worshipped and adored all through Sumerian
history'' \textsuperscript{\cite[p.71]{kramer}}. When turning to
Hittite literature, we are overwhelmed with texts that describe how to
appease an angry god, the correct way to celebrate festivals, and even
how to appeal to the gods for a cure for an impotent husband.  Many of
the interactions with the gods are formulaic, represented in the
literature almost as recipes for divine intervention.  For example,
Paskuwatti walks the reader through a ritual process as she ``make[s]
offerings to Uliliyassi on his behalf and entreat her (Uliliyassi) for
three days.'' \textsuperscript{\cite[p.2014]{sasson}}.  Another
example of this utility is the line of questions presented during the
interaction with an oracular session, such as ``Since a road accident
was ascertained for His Majesty, is this road accident due to the
anger of some deity?'' \textsuperscript{\cite[p.30]{beal}}.  Questions
were even repeated either to verify the answer, or in an attempt for a
more favorable one.
%% ~~~
\begin{quote}
Will there not be an epidemic up in Hattusa and will a general sudden
death not make us flee down from Hattusa? (Answer: Yes, there will
be.) We repeat the question. (Answer: No, there won't
be.) \textsuperscript{\cite[p.30]{beal}}.
\end{quote}
%% ~~~
We see that Hittite literature represents the gods more as
supernatural humans who hold a divine office than Sumerian literature.
Sumerian literature is not devoid of such a \emph{quid pro quo}
nature, but the main distinction is that the Hittites leave ``no trace
of any spontaneous prayer or of any prayer that was purely
adorative.''  \textsuperscript{\cite[p.359]{johnston}}.  The Hittite
gods share many anthropomorphic characteristics with those of the
Sumerians and even directly borrow gods such as An/Anu from the
Sumerian pantheon, but their role diverged at some point from the
Sumerian subjects of adoration to the supernatural beings who hold a
divine office.







%% \newpage
\bib

\end{document}



